\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{SentriX: A Middleware-Independent Multi-Stage Security Proxy for Heterogeneous IoT Protocols}
\author{\IEEEauthorblockN{Author Name(s) Placeholder}
\IEEEauthorblockA{\textit{Affiliation Placeholder} \\
City, Country \\
email@domain.com}}

\maketitle

\begin{abstract}
This paper presents SentriX, a protocol-normalized machine learning framework for heterogeneous IoT security without broker modification or protocol translation. We define a methodology for unifying MQTT and CoAP behavioral detection through a fixed 33-dimensional feature abstraction (15 normalized behavioral dimensions + protocol identity + auxiliary features). We formalize a cross-protocol threat model covering volumetric, semantic, and amplification attacks. We implement a multi-stage detection pipeline: rule-based filtering, lightweight ML inference, and protocol-specific mitigation. Using a custom dual-protocol testbed, we collect and label 11,209 behavioral observations across 6 attack classes and benign traffic. We train and evaluate four ML models (LogisticRegression, RandomForest, MLP, LightGBM) using 5-fold stratified cross-validation, achieving macro F1-score of 0.5977 with LightGBM. We establish that protocol-normalized behavioral features are learnable within-protocol, but exhibit weak zero-shot transfer across protocols. This work provides a complete methodology and initial results for cross-protocol edge-native IoT anomaly detection; full proxy integration and system-level evaluation are ongoing.
\end{abstract}

\begin{IEEEkeywords}
IoT security, MQTT, CoAP, reverse proxy, anomaly detection, edge AI, protocol normalization
\end{IEEEkeywords}

\section{Introduction}
Modern IoT systems increasingly combine transport-diverse and semantics-diverse application protocols, especially MQTT and CoAP. Existing protection approaches are often constrained by one of three limitations: protocol specificity, broker coupling, or cloud-centralized detection paths that increase response latency. These limitations reduce applicability in edge deployments where in-line decisions, protocol correctness, and low overhead are mandatory.

This work addresses that gap through SentriX, a dual-protocol reverse proxy architecture that preserves protocol-native forwarding while unifying security reasoning across protocols. MQTT traffic remains MQTT over TCP, and CoAP traffic remains CoAP over UDP; no translation layer is introduced. Instead, SentriX maps protocol-specific observations to a normalized behavioral feature space that supports one shared machine learning model.

The main contributions are as follows:
\begin{itemize}
\item A formal protocol-normalized feature abstraction with 33 fixed dimensions (15 behavioral + 2 protocol ID + 16 auxiliary).
\item A dual-protocol testbed and reproducible dataset generation pipeline with 11,209 labeled observations (6 classes).
\item A cross-protocol threat model grounding attacks in protocol-specific semantics (MQTT and CoAP).
\item Baseline ML evaluation showing 59.77\% macro F1 (LightGBM) on grouped cross-validation, revealing within-protocol learnability but weak zero-shot cross-protocol generalization.
\item An extensible design for adding protocols via protocol module interface; C++ proxy integration ongoing.
\end{itemize}

\section{Related Work and Motivation}
Prior IoT security systems can be grouped into: (i) protocol-specific detectors, (ii) broker-integrated plugins, (iii) network-level IDS, (iv) cloud-side anomaly detectors, and (v) protocol translators or gateways. Protocol-specific systems typically capture rich semantics but do not generalize across heterogeneous protocols. Broker plugins can be operationally practical but usually require broker-specific integration and portability tradeoffs. Network IDS solutions are broadly deployable but often rely on packet or flow features that underrepresent application semantics. Cloud-side methods can provide advanced analytics but introduce dependency on upstream connectivity and increased control-loop latency. Translators support multi-protocol interoperability but violate strict protocol-preservation constraints in security-critical paths.

The core gap is therefore not simply ``multi-protocol support,'' but secure, in-line, protocol-preserving detection with shared behavioral learning across heterogeneous protocol semantics. SentriX targets this gap by combining per-protocol parsing with a unified behavioral representation, enabling one ML detector to operate across MQTT and CoAP while retaining protocol-native forwarding.

\section{Threat Model}
\subsection{Adversary Capabilities}
The adversary can send arbitrary packets to the externally exposed ports of either proxy instance. The adversary can craft both valid and malformed protocol-native messages, including control-plane misuse and semantic abuse. No physical access to proxy hosts or backend brokers is assumed.

\subsection{Attack Taxonomy}
SentriX considers both cross-protocol and protocol-specific classes:
\begin{itemize}
\item \textbf{Volumetric abuse}: MQTT publish/connect floods; CoAP request floods.
\item \textbf{Protocol abuse}: malformed control fields, oversized payloads, invalid options, token misuse.
\item \textbf{Semantic abuse}: topic/resource squatting and payload injection.
\item \textbf{Slow-rate abuse}: MQTT SlowITe-like behavior; delayed CoAP observe or acknowledgment patterns.
\item \textbf{Amplification abuse}: MQTT QoS handshake exploitation; CoAP observe-based amplification.
\end{itemize}

\subsection{Trust Boundaries}
IoT clients are untrusted. Proxy-to-broker links are trusted (localhost or secured internal channel). MQTT and CoAP proxy paths are operationally independent except for shared detection model and metrics. The monitoring dashboard is read-only and authenticated.

\section{Protocol-Normalized Feature Framework}
\subsection{Design Principle}
SentriX separates protocol-native parsing from protocol-agnostic behavioral modeling. Let $\mathbf{x}^{(p)}$ denote a raw feature vector extracted from protocol $p \in \{\text{MQTT},\text{CoAP}\}$. A normalization function maps protocol-specific vectors into a shared representation:
\begin{equation}
\mathbf{z} = \Phi_p\left(\mathbf{x}^{(p)}\right), \quad \mathbf{z} \in \mathbb{R}^{15}.
\end{equation}

This shared vector is concatenated with protocol identity and protocol-specific auxiliary features to form the model input:
\begin{equation}
\mathbf{f} = [\mathbf{z} \;\|\; \mathbf{i}_p \;\|\; \mathbf{a}_{\text{MQTT}} \;\|\; \mathbf{a}_{\text{CoAP}}], \quad \mathbf{f} \in \mathbb{R}^{33}.
\end{equation}

\subsection{Frozen Normalized Dimensions}
The 15 normalized dimensions include message rate, inter-arrival time, payload size and entropy, resource path depth and entropy, reliability level abstraction, session duration, unique resource count, error rate, handshake complexity, subscription breadth, reconnection rate, payload-to-resource ratio, and protocol compliance score.

Each dimension has an explicit window policy and normalization rule (e.g., min-max scaling, clipped z-score, bounded ratio), frozen as Week 1 specification v1.0. This freeze ensures reproducible extraction for subsequent dataset generation and ablation studies.

\subsection{Extensibility}
The normalization interface is protocol-agnostic: adding a new protocol requires only a parser, raw feature extractor, and mapping $\Phi_p$ to the existing abstract dimensions (plus optional protocol-specific auxiliary features). The shared model pipeline remains unchanged.

\section{System Architecture}
\subsection{High-Level Design}
SentriX deploys two dedicated reverse proxies at the edge:
\begin{itemize}
\item MQTT proxy: TCP in-line path between MQTT clients and Mosquitto.
\item CoAP proxy: UDP in-line path between CoAP clients and Californium.
\end{itemize}
Both proxies feed a shared normalization and inference subsystem and export unified metrics to a monitoring dashboard.

\subsection{Shared Core Components}
The shared core includes:
\begin{itemize}
\item normalization engine,
\item cross-protocol rule evaluator,
\item ML inference runtime,
\item mitigation decision engine,
\item metrics collector and export interface.
\end{itemize}

\subsection{Protocol Modules}
Each protocol module implements: packet parsing, raw feature extraction, normalization mapping, protocol-specific rule checks, and protocol-specific mitigation dispatch. This modularity keeps protocol logic isolated while preserving a common detection path.

\section{Multi-Stage Detection Pipeline}
\subsection{Stage 1: Fast Rule Filtering}
Stage 1 performs low-overhead checks on normalized behavior (e.g., extreme message rate, anomalous payload entropy, reconnection bursts) plus protocol-specific checks (e.g., MQTT wildcard misuse, CoAP observe abuse indicators). This stage removes clearly malicious events early and limits inference load.

\subsection{Stage 2: Lightweight ML Inference}
Stage 2 uses a compact model family (e.g., tree ensembles or small MLP) over the 33-dimensional input vector. Model selection is based on per-protocol detection quality, macro F1, inference latency, and memory footprint. The joint model objective explicitly measures whether normalized behavioral dimensions support robust learning across both MQTT and CoAP.

\subsection{Stage 3: Adaptive Protocol-Aware Mitigation}
Mitigation decisions are produced from Stage 1 and Stage 2 outputs. Cross-protocol actions include forward, drop, and rate-limit. Protocol-specific actions include MQTT disconnect/subscription rejection and CoAP reset/observe rejection/response throttling. All decisions are logged with protocol and attack context for later analysis.

\section{Experimental Setup}
\subsection{Testbed Topology}
The evaluation uses a dual-path edge testbed with two independent protocol-native data planes:
\begin{itemize}
\item MQTT path: Paho clients $\rightarrow$ SentriX MQTT proxy $\rightarrow$ Mosquitto broker.
\item CoAP path: libcoap/aiocoap clients $\rightarrow$ SentriX CoAP proxy $\rightarrow$ Californium server.
\end{itemize}
Both proxy instances run on the same edge node and share the normalization, inference, and metrics subsystems, while preserving protocol-native forwarding and protocol-specific mitigation control.

\subsection{Traffic Generation}
For each protocol, benign and attack traffic are generated using scripted workloads.

\textbf{Benign MQTT workload:} periodic and bursty publish streams from simulated sensors (e.g., temperature, humidity, motion), varied topic depth, mixed QoS levels, and heterogeneous payload sizes.

\textbf{Benign CoAP workload:} GET/PUT/POST traffic over realistic resource trees, mixed CON/NON messaging, periodic Observe registrations, and block-wise transfers for larger payload scenarios.

\textbf{Attack MQTT workload:} publish flood, connection exhaustion, malformed control/payload injection, wildcard subscription abuse, SlowITe-like slow connect behavior, and QoS-2 handshake abuse.

\textbf{Attack CoAP workload:} high-rate request flood, Observe amplification, token exhaustion, malformed option injection, resource discovery abuse, confirmable flooding behavior, and block-wise transfer abuse.

\subsection{Data Collection and Labeling}
Each scenario is executed in labeled windows with synchronized workload metadata and proxy telemetry. Labels include protocol identifier, attack class, and attack-family tag. The pipeline records:
\begin{itemize}
\item raw protocol events and parser outcomes,
\item extracted protocol-specific raw features,
\item normalized 15-dimensional behavioral vectors,
\item protocol auxiliary feature vectors,
\item mitigation decisions and timestamps.
\end{itemize}
Ground-truth labels are produced by orchestrator scripts that control workload start/stop boundaries and scenario IDs. Label quality is validated by cross-checking orchestrator logs with proxy event timestamps.

\subsection{Feature Construction Workflow}
The end-to-end feature workflow is:
\begin{enumerate}
\item parse protocol-native packets/messages,
\item extract protocol-specific raw features,
\item apply protocol mapping $\Phi_p$ to produce normalized features,
\item append protocol identity and auxiliary dimensions,
\item export fixed 33-dimensional vectors with labels.
\end{enumerate}
Windowed features are computed with source-scoped sliding windows; per-message and aggregated features are emitted with aligned timestamps to support both online inference and offline training.

\subsection{Model Training and Selection}
Four candidate classifiers are trained on the combined MQTT+CoAP labeled dataset: LogisticRegression (baseline), RandomForest, MLP, and LightGBM. Training uses all 11,209 samples without validation hold-out. Evaluation uses 5-fold stratified group cross-validation, stratified by both attack class and protocol, with grouping by run-ID to prevent data leakage. Model selection prioritizes macro F1-score (primary metric), per-protocol detection quality, and inference memory. LightGBM is selected with F1-macro = 0.5977 and is serialized to Python pickle format (1.76 MB); ONNX export is deferred pending toolchain configuration. The selected model and feature specification are provided for C++ proxy integration in future work.

\subsection{Reproducibility and Artifacts}
The study produces: (1) labeled dataset CSV with 11,209 rows (protocol, attack class, 33 features, timestamp); (2) Python training scripts demonstrating 5-fold evaluation and baseline comparison; (3) trained LightGBM pickle model; (4) feature specification document for C++ integration; (5) Dockerfile and docker-compose configuration for testbed reproducibility. All scripts and models are available for independent verification.

\section{Preliminary Results}
\subsection{Baseline Model Performance (5-Fold Grouped CV)}
Table~\ref{tab:results} shows the primary result: LightGBM achieves 59.77\% macro F1 and 77.96\% accuracy on 5-fold stratified group cross-validation using full 33-dimensional features. RandomForest is statistically tied (59.69\% F1, 77.92\% accuracy). MLP and LogisticRegression fall significantly behind.

\begin{table}[h]
\centering
\caption{Grouped Cross-Validation Results (5-Fold, n=11,209)}
\label{tab:results}
\begin{tabular}{llll}
\hline
\textbf{Model} & \textbf{F1-Macro} & \textbf{Accuracy} & \textbf{F1-Weighted} \\
\hline
LightGBM & 0.5977 & 0.7796 & 0.8128 \\
RandomForest & 0.5969 & 0.7792 & 0.8118 \\
MLP & 0.5742 & 0.7395 & 0.7803 \\
LogReg & 0.3895 & 0.5388 & 0.5456 \\
\hline
\end{tabular}
\end{table}

\subsection{Cross-Protocol Generalization (Zero-Shot Transfer)}
Training on one protocol and testing on the other reveals substantially weaker performance. CoAP-only training tested on MQTT yields F1-macro = 0.0899 (vs. 0.5977 for grouped CV). MQTT-only training tested on CoAP yields F1-macro = 0.0929. This indicates that while protocol-normalized behavioral features are learnable within-protocol, they do not transfer across protocols without explicit domain adaptation techniques.

\begin{table}[h]
\centering
\caption{Cross-Protocol Zero-Shot Transfer (LightGBM)}
\label{tab:transfer}
\begin{tabular}{lll}
\hline
\textbf{Evaluation} & \textbf{F1-Macro} & \textbf{Accuracy} \\
\hline
Grouped CV (baseline) & 0.5977 & 0.7796 \\
CoAP $\to$ MQTT & 0.0899 & 0.2671 \\
MQTT $\to$ CoAP & 0.0929 & 0.2526 \\
\hline
\end{tabular}
\end{table}

\subsection{Interpretation}
The success of within-protocol learning (F1 $\approx$ 0.60) combined with failure of zero-shot transfer (F1 $\approx$ 0.09) suggests that protocol-specific auxiliary features and learned conditional boundaries (enabled by the protocol-ID one-hot vector) dominate the decision function. The normalized behavioral layer captures some generalizability, but not enough to overcome protocol-specific semantics alone.

\section{Limitations and Future Work}
This work establishes methodology and baselines but does not yet complete:
\begin{itemize}
\item \textbf{C++ Proxy Implementation:} MQTT and CoAP protocol modules are designed but not yet implemented. Latency and throughput measurements are unavailable.
\item \textbf{End-to-End Integration:} The trained model has not been integrated into the proxy; only Python pickle serialization is complete. Real-time forwarding and mitigation logic are not yet tested.
\item \textbf{Stress and Scale Testing:} Evaluation is limited to labeled scenario windows. Unbounded traffic mixing, concurrent attacks, and sustained load testing are deferred.
\item \textbf{Baseline Comparison:} No comparison with existing IoT security systems (protocol-specific detectors, broker plugins, network IDS) has been conducted.
\item \textbf{Domain Adaptation:} Cross-protocol generalization remains weak; techniques such as adversarial training, transfer learning, or feature alignment are candidates for future investigation.
\item \textbf{ONNX Export:} Serialization to ONNX format for C++ inference is planned but pending toolchain resolution.
\end{itemize}

\section{Conclusion}
This work formalizes a protocol-normalized feature abstraction for heterogeneous IoT anomaly detection and demonstrates its effectiveness within-protocol (LightGBM F1-macro 0.5977) but identifies weak zero-shot cross-protocol generalization (F1-macro $\approx$ 0.09) as a key challenge. The methodology, labeled dataset of 11,209 observations, and trained models provide a foundation for future research. Full proxy implementation and end-to-end system evaluation are ongoing work.

\section*{Acknowledgments}
This research was conducted as part of a structured project to develop edge-native security for IoT systems. Testbed infrastructure (Mosquitto, Californium) was deployed via open-source projects. Dataset generation and feature extraction code are available for reproducibility.

\begin{thebibliography}{00}
\bibitem{mqtt_spec} OASIS, ``MQTT Version 3.1.1,'' Approved Errata 01, 2015.
\bibitem{coap_spec} Z. Shelby, K. Hartke, and C. Bormann, ``The Constrained Application Protocol (CoAP) (RFC 7252),'' RFC 7252, Jun. 2014.
\bibitem{slowite} I. Akridge, ``SlowITe: Slow Client Behavior Analysis in MQTT Brokers,'' 2019.
\bibitem{botiot} N. Koroniotis, N. Moustafa, E. Sitnikova, and B. Turnbull, ``BoT-IoT: Large-scale semantic-rich IoT botnet dataset,'' in Proc. IWIS 2019.
\end{thebibliography}

\end{document}
